---
title: "Network Analysis with R"
author: Yue Hu
output: html_notebook
---

Before the data exploration, here are commonly used data input, management, and network analysis packages.
We need to load them for later useage. 

```{r setup, include = TRUE}
# chunk property setting
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# package loading
library(pacman)
p_load(ggplot2, GGally, # Visualization
       network, sna, igraph,# Applied
       readxl, haven, descr, stringi, stringr, arm, car, stargazer, data.table, broom, tibble, tidyr, purrr, dplyr) # data wrangling

# replication
set.seed(313)
```

Source: https://briatte.github.io/ggnet/

# Network Data
## Data Format
Network data are used to record network relations including nodes and edges.
Commonly, such data can be categorized in three dimensions: 

* Data format: adjacency vs. edgelist.
* Edge weight: weighted vs. unweighted.
* Edge direction: directed vs. nondirected.


Let's first have a look at what do each looks like.
Here we use the `sna::rgraph` function to create toy network data for illustration.

```{r}
# Unweighted nondirected adjacency network ####
net_una <- rgraph(5, mode = "graph", tprob = 0.5)
net_una

# Unweighted nondirected edgelist ####
net_une <- rgraph(5, mode = "graph", tprob = 0.5, return.as.edgelist = TRUE)
net_une

# weighted directed edgelist ####
net_wde <- rgraph(5, mode = "graph", tprob = 0.5, return.as.edgelist = TRUE)
net_wde[, 3] <- sample(1:5, nrow(net_wde), replace = TRUE)
net_wde

# Unweighted directed adjacency network ####
net_uda <- rgraph(5, mode = "digraph", tprob = 0.5)
net_uda
```

As shown above, the rows and columns of an adjacency network are both indices of nodes, and the cells represent edge status between the nodes. 
The corresponding edgelist records the same information but in a dyadic way, where the data are recored in three columns: the indices of the first node, the indices of the second node, and the edge status.
Regarding the edge status, a unweighted network, 1 represents the nodes are connected, while 0 represent they are not.
When the network is weighted, instead of 1/0, the edge status shows a numeric measure of the strength of the connections between nodes.
Finally, in nondirected network, the connections between nodes are always symmetric. That is, if A connects with B (A->B = 1), B is then definitely connected with A (B->A = 1).
In a (unweigthed) direct network, this restriction is released. When A->B = 1, B->A can be either 1 or 0.

NOTE: Some network analyses in R may require not only a matrix of the network relations. The matrix should also be saved as a `network` project.
In such case, one more step after getting the correct network matrix: 

```{r}
net_unaNET <- network(net_una, directed = FALSE)
class(net_unaNET)
```


## Data converting
Although the same network can be recorded in adjacency matrix and edgelist, some analysis may require a specific type of data format.
Then, how do we convert them between each other for the sake of analysis?


```{r}
# Adjacency to Edgelist ####
net_una2e <- as.edgelist(net_unaNET)
net_una2e

# Edgelist to Adjacency ####
net_une2a <- network(net_une, directed = FALSE) %>% 
  # "%>%" is a pipe sign from the `dplyr` package meaning taking the result of the previous step to the following step as the first argument of the function.
  as.sociomatrix()
net_une2a
```

## Longitudinal Network
In this section, we discuss a special demand of data management: longitudinal network.
It refers to the data recording not only the spatial distribution of a network but also the time trend. 
For preparing analyses for such data, there are generally two ways to manage the data, in data matrix or list. 

### Longitudinal network data matrix
This type of data can be directly used in `tnet` functions for TERGM analyses.
The format is merely adding the time line in an edgelist:
```{r}
net_unt <- rgraph(5, mode = "graph", tprob = 0.5, return.as.edgelist = TRUE) %>%
  as.data.frame() %>%
    mutate(t = sample(1900:2017, nrow(.)), 
           t = as.character(t), 
           t = as.POSIXct(t, format = "%Y")) %>%
             select(-V1, -V2, -V3, everything()) %>% # reorder the columns
               rename(i = V1, j = V2, w = V3) # unnecessary step; just for showing the nature of the variables

net_unt
```

Notice: 

1. For a longitudinal data for `tnet` functions, there is a special requirement of the order of the columns. 
The first column should be the time trend, and then the indices of the first, second node, and finally the weight of edges (weighting 1 for unweighted network).

2. The time trend has to be coded as `POSIXct` classes (including `Date`). 

### Temporial list of Network 
Some packages (e.g., `tsna`) may require data to be coded as temporial lists, that is, storing an adjacency network in each element of a list according to time.^[See analytical examples with `tsna` in http://statnet.csde.washington.edu/workshops/SUNBELT/current/ndtv/ndtv_workshop.html]
How can you get data in such format from a dyadic data frame?
Here's an example. 
We create a list of network (which is actually the format we finally need) first and convert it into a dyadic format for later practice. 
Then, using the functions in `purrr` to convert it back to a temporal list of networks.
```{r}
# Creating the toy data ####
net_unt2 <- rgraph(5, m = 3, mode = "graph", tprob = 0.5) %>%
  # gives an array with the first dimension as the network index
  array_tree(margin = 1) 

net_unt2

# Convert to the edgelist ####
net_unt3 <- map_df(net_unt2, function(net){
  t <- sample(1900:2017, 1) %>%
    as.character() %>%
      as.POSIXct(format = "%Y")
  
  net <- as.network(net) %>%
    as.edgelist() %>%
      as.data.frame()
  
  net$t <- t
  return(net)
  })

net_unt3

# Convert back to temporal list of network projects####
net_unt3 <- net_unt3 %>%
  split(.$t) %>%
    map(function(net){
      net <- select(net, -t) %>%
        as.network(directed = FALSE) %>%
          as.sociomatrix})

net_unt3
```

