---
title: "Geographic Data Visualization"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(maps)
library(mapdata)
library(maptools)
library(tidyverse)

```

**"Hello, R!" workshop series V**   
*Iowa Social Science Research Center (ISRC)*

## Preface

* R Basics
* Data Manipulation with R and STATA
* Quantitative Analysis
* Basic Data Visualization
* **Geographic Data Visualization**
* Multilevel Model with R and STATA
* R Professionalization

<br>

Register other workshops in the [ISRC website](http://ppc.uiowa.edu/node/3608).


## Today's Plan

* Simple steps of creating maps with R
* A intro to R map with `ggplot`
* A intro to `ggmap`
* ArcGIS basics for map building

## Map with R

* Draw the world map with only two commands:

```{r world, exercise = TRUE}
maps::map("world", 
    fill = TRUE, # whether fill the regions
    col = rainbow(20), # number of colors to be used
    mar = c(0, 0, 0, 0)) # margin setting
```

## A Map of the United States

```{r us, exercise = TRUE}
maps::map("state", fill = TRUE, col = rainbow(20), mar = c(0, 0, 2, 0))
```

```{r us-solution}
maps::map("state", region = c("iowa", "missouri", "illinois"), fill = TRUE, col = rainbow(3))

maps::map("county", region = "iowa", fill = TRUE, col = rainbow(20))
```

`maps` package includes maps of Canada, France, Italy, New Zealand, and various types of maps of the U.S.

## A Map of China

```{r chinaRaw, exercise = TRUE}
library(mapdata) # including China, Japan, and Pacific Centric World
maps::map("china", fill = FALSE, col = "gray40")
```
